---
title: "`r params$title_override`"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  word_document:
    toc: true
    toc_depth: 2
params:
  data_file: NULL
  spec_file: NULL
  r_dir: NULL
  title_override: "Cohort Profile"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE,
  fig.width = 6.5, fig.height = 4.2, dpi = 200
)

# The report renders once per output format, and Fisher's exact test uses
# simulated p-values. Without a fixed seed the DOCX and HTML versions would
# carry slightly different numbers — exactly the disagreement between
# formats this project promises cannot happen.
set.seed(20260630)

# The report is rendered in a fresh environment, so bring the analysis code in
# explicitly rather than relying on anything being attached already.
if (!is.null(params$r_dir)) {
  for (f in list.files(params$r_dir, pattern = "\\.R$", full.names = TRUE)) {
    source(f, local = TRUE)
  }
}

spec <- read_spec(params$spec_file)
raw  <- read_export(params$data_file)

validation <- validate_export(raw, spec)
stop_if_invalid(validation)

analysis <- build_analysis_data(raw, spec, validation)
quality  <- data_quality(analysis)

y_name <- spec$outcome$name
outcome_tab <- table(analysis[[y_name]])
n_analytic  <- nrow(analysis)

# Only domains that actually have data become sections.
live_domains <- Filter(function(dm) length(domain_vars(analysis, spec, dm)) > 0,
                       spec$domains)
```

```{r title-block, results = "asis"}
cat("# Overview\n\n")
```

This report describes **`r n_analytic` participants** from the
`r spec$study` study and tests each variable one at a time against the
outcome, **`r spec$outcome$label`**.

The upload contained `r validation$n_rows` rows. After keeping only
`r spec$filter$event_value` records with an outcome result,
`r n_analytic` participants remain.

```{r overview-outcome}
ov <- data.frame(
  Group = names(outcome_tab),
  N = as.integer(outcome_tab),
  `%` = sprintf("%.1f%%", 100 * as.integer(outcome_tab) / n_analytic),
  check.names = FALSE, stringsAsFactors = FALSE
)
style_table(flextable::flextable(ov), title = "Outcome distribution")
```

Every association below is **unadjusted**. Multivariable modelling is
deliberately out of scope for this report.

# Data quality

```{r dq-intro, results = "asis"}
issues <- c(
  length(quality$soft),
  if (!is.null(quality$out_of_range)) nrow(quality$out_of_range) else 0,
  if (!is.null(quality$bad_levels)) nrow(quality$bad_levels) else 0,
  if (!is.null(quality$high_missing)) nrow(quality$high_missing) else 0
)
if (sum(issues) == 0) {
  cat("No data-quality problems were found. Every variable in the spec matched",
      "a column, all values were inside their declared ranges, and every",
      "response was one the spec declares.\n\n")
} else {
  cat("The export passed the checks needed to run the analysis. The points",
      "below did not stop the report but are worth knowing about when reading",
      "it.\n\n")
}
```

```{r dq-soft, results = "asis"}
if (length(quality$soft)) {
  cat("## Columns the spec expected but did not find\n\n")
  for (f in quality$soft) cat("- ", f$message, "\n", sep = "")
  cat("\n")
}
```

```{r dq-range, results = "asis"}
ft <- range_flextable(quality)
if (!is.null(ft)) {
  cat("## Values outside the plausible range\n\n")
  flextable::flextable_to_rmd(ft)
  cat("\n")
}
```

```{r dq-levels, results = "asis"}
ft <- levels_flextable(quality)
if (!is.null(ft)) {
  cat("## Responses the spec does not declare\n\n")
  cat("These were left as missing. They are never folded into an existing",
      "category, because that would hide a change in how the question is",
      "coded.\n\n")
  flextable::flextable_to_rmd(ft)
  cat("\n")
}
```

```{r dq-highmiss, results = "asis"}
hm <- quality$high_missing
if (!is.null(hm) && nrow(hm)) {
  cat("## Variables above the missingness threshold\n\n")
  cat(sprintf("More than %.0f%% missing. Summaries for these are based on very few participants.\n\n",
              100 * spec$options$missingness_threshold))
  for (i in seq_len(nrow(hm))) {
    cat(sprintf("- %s: %.1f%% missing\n", hm$label[i], hm$pct_missing[i]))
  }
  cat("\n")
}
```

# Descriptive tables by outcome status

```{r descriptives, results = "asis"}
letters_seq <- letters[seq_along(live_domains)]
for (i in seq_along(live_domains)) {
  dm <- live_domains[[i]]
  ft <- descriptive_flextable(
    analysis, spec, dm,
    title = sprintf("Table 1%s. %s by %s status", letters_seq[i], dm,
                    outcome_positive_label(spec)))
  if (is.null(ft)) next
  cat("\n## ", dm, "\n\n", sep = "")
  flextable::flextable_to_rmd(ft)
  cat("\n")
}
```

# Missingness

## Per-variable missingness

```{r missingness}
missingness_flextable(quality, spec)
```

## Missingness mechanism

```{r mcar, results = "asis"}
mc <- mcar_result(analysis, spec)
if (isTRUE(mc$ok)) {
  cat(sprintf(
    paste("Little's test for data missing completely at random was run over the",
          "%d numeric variables below the %.0f%% missingness threshold.",
          "The statistic is %.1f on %d degrees of freedom (p = %s), across %d",
          "missingness patterns.\n\n"),
    mc$n_vars, 100 * spec$options$missingness_threshold,
    mc$statistic, mc$df, fmt_p(mc$p.value), mc$missing.patterns))
  if (length(mc$dropped)) {
    cat(sprintf(paste("%s %s left out of this test because %s an exact",
                      "combination of other variables.\n\n"),
                paste(mc$dropped, collapse = ", "),
                if (length(mc$dropped) == 1) "was" else "were",
                if (length(mc$dropped) == 1) "it is" else "they are"))
  }
  cat(if (!is.na(mc$p.value) && mc$p.value < 0.05) {
    paste("A small p-value argues against missing completely at random, so",
          "a complete-case analysis may be biased.\n\n")
  } else {
    paste("There is no strong evidence against missing completely at random",
          "among these variables.\n\n")
  })
} else {
  cat(mc$message, "\n\n")
}
```

```{r missing-vs-outcome}
mv <- missing_vs_outcome(analysis, spec)
missing_vs_outcome_flextable(mv)
```

# Variable consistency checks

```{r checks, results = "asis"}
ft <- checks_flextable(quality)
if (is.null(ft)) {
  cat("The spec declares no consistency rules.\n\n")
} else {
  cat("Each rule below should hold for a plausible record. Records where the",
      "rule cannot be evaluated, usually because one of the variables is",
      "missing, are excluded from the denominator rather than counted as",
      "passing.\n\n")
  flextable::flextable_to_rmd(ft)
  cat("\n")
}
```

# Plots of key variables versus the outcome

```{r plots, fig.height = 5.2}
uni_all <- univariate_results(analysis, spec, live_domains)
p <- outcome_plots(analysis, spec, uni_all)
if (!is.null(p)) p
```

```{r plots-note, results = "asis"}
if (is.null(p)) cat("No numeric variables were estimable, so there is nothing to plot.\n\n")
```

# Univariate logistic associations

Each variable is entered on its own. Odds ratios are unadjusted, with Wald
95% confidence intervals. A variable is reported as **not estimable** rather
than given a number when it has fewer than
`r spec$options$min_complete_cases` complete cases, has no variation, or
separates the outcome perfectly — a separating variable would otherwise
produce an enormous odds ratio that reads as a finding.

```{r univariate, results = "asis"}
for (dm in live_domains) {
  res <- univariate_results(analysis, spec, dm)
  ft <- univariate_flextable(
    res, title = sprintf("Univariate associations: %s", dm))
  if (is.null(ft)) next
  cat("\n## ", dm, "\n\n", sep = "")
  flextable::flextable_to_rmd(ft)
  cat("\n")
}
```

```{r nonestimable, results = "asis"}
ne <- uni_all[nzchar(uni_all$note), , drop = FALSE]
if (!is.null(ne) && nrow(ne)) {
  cat("## Variables reported as not estimable\n\n")
  for (i in seq_len(nrow(ne))) {
    cat(sprintf("- **%s** — %s\n", ne$label[i], ne$note[i]))
  }
  cat("\n")
}
```

# Notes and caveats

- Every association in this report is **unadjusted**. No variable is
  controlled for any other. Multivariable and penalised models are out of
  scope here and live in a separate pipeline.
- Continuous variables are summarised as median [IQR] and mean (SD) and
  compared with the Wilcoxon rank-sum test, which does not assume normality.
- Categorical variables are compared with Pearson's chi-square, falling back
  to a simulated Fisher's exact test when any expected cell count is below
  five.
- Values outside the ranges declared in the spec were set to missing before
  analysis. The count of such values is in the data-quality section.
- Responses not declared in the spec were treated as missing rather than
  grouped into an existing category.
- p-values are not adjusted for multiple comparisons. With this many tests,
  some results at p < 0.05 are expected by chance alone.

```{r provenance, results = "asis"}
cat(sprintf("\n---\n\nSource file: `%s`  \nSpec: `%s`  \nGenerated: %s\n",
            basename(params$data_file), basename(params$spec_file),
            format(Sys.time(), "%Y-%m-%d %H:%M")))
```
